---
title: "Machine Learning - Block02 Assignment 01"
author: "Agustín Valencia"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mboost)
library(randomForest)
library(partykit)
library(ggplot2)
set.seed(1234567890)
```


# 1. Ensemble Methods
The file spambase.csv contains information about the frequency of various words, characters, etc. for a total of 4601 e-mails. Furthermore, these e-mails have been classified as spams (spam = 1) or regular e-mails (spam = 0). You can find more information about these data at https://archive.ics.uci.edu/ml/datasets/Spambase. 

Your task is to evaluate the performance of Adaboost classification trees and random forests on the spam data. Specifically, provide a plot showing the error rates when the number of trees considered are 10, 20, . . . , 100. To estimate the error rates, use 2/3 of the data for training and 1/3 as hold-out test data.

To learn Adaboost classification trees, use the function blackboost() of the R package mboost. Specify the loss function corresponding to Adaboost with the parameter family. To learn random forests, use the function randomForest of the R package randomForest. To load the data, you may want to use the following code:

```{r ensemble}
sp <- read.csv2("data/spambase.csv")
sp$Spam <- as.factor(sp$Spam)
```

## Solution

For trees trained using adaboost we have:
```{r}
## Splitting data
n <- dim(sp)[1]
idxs <- sample(1:n, floor(2*n/3))
train <- sp[idxs,]
test <- sp[-idxs,]

get_missed <- function (true, predicted) {
    confussion <- table(true, predicted)
    #tn <- confussion[1,1]
    #tp <- confussion[2,2]
    fn <- confussion[1,2]
    fp <- confussion[2,1]
    total <- sum(confussion)
    #success <- (tp + tn) / total * 100
    miss <- (fp + fn) / total * 100
    return(miss)
}

# Training
nums <- seq(10,100,10)
formula <- Spam ~ .
error_rates_train_bb <- c()
error_rates_test_bb <- c()
error_rates_train_rf <- c()
error_rates_test_rf <- c()
depths <- c()
for (i in nums) {
    bb <- blackboost (
            Spam ~ ., 
            data = train, 
            family = AdaExp(), 
            control = boost_control(mstop = i)
         )
    
    rf <- randomForest(
            Spam ~ .,
            data = train, 
            ntree = i
    )
    
    predicted <- predict(bb, train, type = "class")
    miss <- get_missed(train$Spam, predicted)
    error_rates_train_bb <- append(error_rates_train_bb, miss)
    
    predicted <- predict(bb, test, type = "class")
    miss <- get_missed(test$Spam, predicted)
    error_rates_test_bb <- append(error_rates_test_bb, miss)
    
    predicted <- predict(rf, train, type = "class")
    miss <- get_missed(train$Spam, predicted)
    error_rates_train_rf <- append(error_rates_train_rf, miss)
    
    predicted <- predict(rf, test, type = "class")
    miss <- get_missed(test$Spam, predicted)
    error_rates_test_rf <- append(error_rates_test_rf, miss)
       
    depths <- append(depths, i)
}

p <- ggplot()
p <- p + geom_line(aes(x = depths, y = error_rates_train_bb), color = "red")
p <- p + geom_line(aes(x = depths, y = error_rates_test_bb), color = "blue")
p <- p + geom_line(aes(x = depths, y = error_rates_train_rf), color = "black")
p <- p + geom_line(aes(x = depths, y = error_rates_test_rf), color = "green")
p <- p + geom_point(aes(x = depths, y = error_rates_train_bb), color = "red")
p <- p + geom_point(aes(x = depths, y = error_rates_test_bb), color = "blue")
p <- p + geom_point(aes(x = depths, y = error_rates_train_rf), color = "black")
p <- p + geom_point(aes(x = depths, y = error_rates_test_rf), color = "green")

p
```

# 2. Mixture Models

Your task is to implement the EM algorithm for mixtures of multivariate Benoulli distributions. Please use the template in the next page to solve the assignment. Then, use your implementation to show what happens when your mixture models has too few and too many components, i.e. set $K = 2, 3, 4$ and compare results. Please provide a short explanation as well.


```{r}
max_it <- 100 # max number of EM iterations
max_it <- 10 # max number of EM iterations
min_change <- 0.1 # min change in log likelihood between two consecutive EM iterations
N=1000 # number of training points
D=10 # number of dimensions
x <- matrix(nrow=N, ncol=D) # training data

# [AV] Step 1: Set pi and mu to some initial values.
true_pi <- vector(length = 3) # true mixing coefficients
true_mu <- matrix(nrow=3, ncol=D) # true conditional distributions
true_pi=c(1/3, 1/3, 1/3)
true_mu[1,]=c(0.5,0.6,0.4,0.7,0.3,0.8,0.2,0.9,0.1,1)
true_mu[2,]=c(0.5,0.4,0.6,0.3,0.7,0.2,0.8,0.1,0.9,0)
true_mu[3,]=c(0.5,0.5,0.5,0.5,0.5,0.5,0.5,0.5,0.5,0.5)
plot(true_mu[1,], type="o", col="blue", ylim=c(0,1))
points(true_mu[2,], type="o", col="red")
points(true_mu[3,], type="o", col="green")

# Producing the training data
for(n in 1:N) {
    k <- sample(1:3,1,prob=true_pi)
    for(d in 1:D) {
        x[n,d] <- rbinom(1,1,true_mu[k,d])
    }
}
K=3 # number of guessed components
z <- matrix(nrow=N, ncol=K) # fractional component assignments
pi <- vector(length = K) # mixing coefficients
mu <- matrix(nrow=K, ncol=D) # conditional distributions
llik <- vector(length = max_it) # log likelihood of the EM iterations
# Random initialization of the paramters
pi <- runif(K,0.49,0.51)
pi <- pi / sum(pi)
for(k in 1:K) {
    mu[k,] <- runif(D,0.49,0.51)
}
pi
mu

for(it in 1:max_it) {
    plot(mu[1,], type="o", col="blue", ylim=c(0,1))
    points(mu[2,], type="o", col="red")
    points(mu[3,], type="o", col="green")
    # points(mu[4,], type="o", col="yellow")
    # E-step: Computation of the fractional component assignments
    # Your code here
    aux <- log(sum())
    llik[i] <- sum()
    # Log likelihood computation.
    # Your code here
    cat("iteration: ", it, "log likelihood: ", llik[it], "\n")
    flush.console()
    # Stop if the lok likelihood has not changed significantly
    # Your code here
    # M-step: ML parameter estimation from the data and fractional component assignments
    # Your code here
}
pi
mu
plot(llik[1:it], type="o")
```


